Modeling Intervention Strategies for United States TB Control

### 2013-08-20 in_progress/presentations/paper/draft.tex
added background to paper

\documentclass{amsart}

\title{INSERT TITLE HERE}
\begin{document}
\maketitle

\section{Introduction}
Epidemiological models allow public health professionals to predict and analyze
disease dynamics and intervention effectiveness. The most common examples of
such models are compartmental differential equation models, in which the
population is split between several possible health states, with flow between
each state given according to deterministic differential equations. In 2012, 
Hill, Becerra, and Castro implemented a compartmental
differential equaiton model of tuberculosis (TB) in the United States (US).
Their model utilized five health states and two subpopulations, US born (USB)
and foreign born (FB) for a total of 10 compartments. They used this model to
evaluate several possible intervention strategies, and ultimately conclude that
though increasing LTBI treatment was a good intervention strategy, the US was
unlikely to meet their stated goal of elimination of TB in the US by 2100. In
this work, the Hill model was extended in several key ways. First, additional
tracking capabilities were added to the Hill model, such that it can now report
further granularity in the disease dynamics. Further, economic components were
added to the model in order to project the US health care system (HCS) costs due
to TB given our current policy as well as for various interventions. Finally, a
population level, agent based implementation of the Hill Model was created, in
order to validate the Hill Model against the natural stochasticity present in
real world disease spread.

@@

\subsection{Basic Implementation}
Both the basic Hill Model and this extended Hill Model were implemented in
\texttt{R} as a system of differential equations, which were solved via the
$\texttt{lsoda}$ routine. The systems of differential equations used in the Basic
Hill Model and the extended Hill Model are shown below in Figure 1 and Figure 2
respectively.

INSERT FIGURES HERE

>>>>

\documentclass{amsart}
\usepackage[pdftex]{graphicx}

\title{Modeling Intervention Strategies for United States TB Control}
\begin{document}
\maketitle

\section{Introduction}
Epidemiological models allow public health professionals to predict and analyze
disease dynamics and intervention effectiveness. The most common examples of
such models are compartmental differential equation models, in which the
population is split between several possible health states, with flow between
each state given according to deterministic differential equations. In 2012, 
Hill, Becerra, and Castro implemented a compartmental
differential equaiton model of tuberculosis (TB) in the United States (US).
Their model utilized five health states and two subpopulations, US-born (USB)
and foreign-born (FB) for a total of 10 compartments. They used this model to
evaluate several possible intervention strategies, and ultimately conclude that
though increasing LTBI treatment was a good intervention strategy, the US was
unlikely to meet their stated goal of elimination of TB in the US by 2100. In
this work, the Hill model was extended in several key ways. First, additional
tracking capabilities were added to the Hill model, such that it can now report
further granularity in the disease dynamics. Further, economic components were
added to the model in order to project the US health care system (HCS) costs due
to TB given our current policy as well as for various interventions. Finally, a
population level, agent based implementation of the Hill Model was created, in
order to validate the Hill Model against the natural stochasticity present in
real world disease spread.

\section{Background}

\subsection{The Hill Model}
A flowchart representation of the Hill Model is shown in Figure 1.  Each compartment
represents a different possible health state with respect to TB for every 
US-born or foreign-born individual, and arrows between different compartments
represent possible transitions between states.  Individuals also leave the model 
from all compartments due to natural death, which is left out of the figure for clarity.  \\

\begin{figure}
\includegraphics[scale=0.25]{figures/HillModelFlowChart.png}
\end{figure}

The majority of USB and FB individuals fall into the Susceptible (S) category, which includes everyone
who is uninfected and has not been exposed to TB.  After exposure to an individual
with TB, a person in the Susceptible compartment can develop Latent TB Infection (LTBI) and changes
health states to either the Acute LTBI (F) or Chronic LTBI (L) compartment.  Latently
infected individuals are not infectious, but have some risk of developing active TB infection
over time.  However, the rates of disease progression are not equal, and individuals in the
Acute LTBI compartment have a higher risk of developing active TB than those in
the Chronic LTBI compartment.  Individuals in the Chronic LTBI compartment
may also be exogenously re-infected and transition to the Acute LTBI compartment.  \\

Latently infected individuals may progress to one of two active TB states: Infectious TB (I) or 
Non-Infectious TB (J).  Individuals in both compartments have an increased risk of death from
active TB infection, but only individuals in the Infectious TB compartment are contagious.  
In addition, individuals in all of the infected compartments (F, L, I, J) may be treated or self-cure
themselves of their respective TB health condition.  However, in the model, treatment or self-cure from TB
does not grant immunity, and all healthy individuals are grouped in the Susceptible compartment
and may be re-infected at a later time.  \\

@@

\subsection{Basic Implementation}
Both the basic Hill Model and this extended Hill Model were implemented in
\texttt{R} as a system of differential equations, which were solved via the
$\texttt{lsoda}$ routine. The systems of differential equations used in the Basic
Hill Model and the extended Hill Model are shown below in Figure 2 and Figure 3
respectively.

\begin{figure}
\includegraphics[scale=0.75]{figures/BasicHillEquations.pdf}
\end{figure}

The vector variables $S_{0}, F_{0}, L_{0}, I_{0}, J_{0}$ contain the number of US-born individuals in the S, F, L, I, J compartments respectively, and $S_{1}, F_{1}, L_{1}, I_{1}, J_{1}$ contain foreign-born individuals.  $N_{0}$ and $N_{1}$ are the total populations of US-born and foreign-born individuals.
The constants $\rho$ and $\alpha$ are birth rates, while $\mu_{i}$, and $\mu_{d}$ are death rates.  
A complete list and descriptions of all constants used in the model can be found in the appendix.
